Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-09 versions/response.v1.txt

> Getting exact regions is what I think the authors should do. The
> approach used to obtain Fig 5 is generally applicable. Explicit,
> relatively simple, expressions are given in the appendix for long-run
> lnls, l_{tau_1,t*}(tau_2,w,x,y). For any given (x*,y*) on a fine grid
> in Figs 2-4, l_{tau_1,t*}(tau_j,w,x,y), can be maximized using
> numerical algorithms over the three parameters w,x,y. Doing so for
> tau_j=tau_1, tau_2 would determine the actual regions of problematic
> estimation corresponding to Figs 2-4. The only potential difficulty
> that I can see is that numerical algorithms might not converge. This
> can either be avoided by considering a cubic grid of starting (w,x,y),
> or through something like the following. For any given set of
> ancestral character state assignments for the distinct set of 8
> patterns 1111, 1110,... (0000 can treated as the same as 1111), the
> long-run lnl is of the form

> l(x,y,w)=a log(1-x) + b log(1+x) + c log(1-y) + d log(1+y) + e log(1-w) + f log(1+w)

> where a,..f>=0 are constants determined by (x*,y*) and what ancestral
> states have been assigned. The function l(x,y,w) has an explicit
> simple maximizer. Thus the overall maximizer is determined as the
> largest maximized l(x,y,w) over all 4^8 possible ancestral state
> assignments: a finite number of comparisons. The actual number of
> possible ancestral state assignments is quite a bit smaller than 4^8
> (for instance 11 is always the ancestral reconstruction for 1111) and
> the authors have a lot of familiarity with what the ancestral
> reconstructions are. I believe Fig 2 can be replaced with an exact (up
> to grid approximation) region of topological inconsistency and Figs
> 3-5 with heat maps of limiting (x,y,w) for tau_1.

Please see above for our response.
%das: I added some stuff below to maybe alleviate some confusion and to show why what this reviewer thinks we should do won't work, but not sure how much is necessary.
%Maybe it adds to why we didn't do this.
It is true that the form given above for l(x,y,w) has an explicit and simple maximizer.
However, the long-run log-likelihood is slightly more complicated than this function, as seen in equation (9).
For both topologies, we have a number of equations of this form that we would need to maximize in a grid search.
The function of the form in l(x,y,w) above was our approximation to the upper and lower bounds of the log-likelihood, and was our choice for these bounds precisely because it has an explicit and simple maximizer.
Doing a grid search over all ancestral states would not be trivial, and we would need to perform numerical optimization for all possible ancestral states to compute the maximum.

@@

%EM I think we could move this to the appendix to throw this reviewer a bone.
%das: does this reviewer mean both the general case and the empirical case or just the empirical case?
%Their review makes me think they take the empirical case the be the "general" case.
%You think we should move both analytic cases to the appendix or just the one?

> Minor Comments:

> -l78-88: Calculations are more natural with `fidelities' and this
> might remain in the appendix but it is much more conventional in the
> literature for x,y,w to be edge-lengths or probabilities of a
> different character states at the tips of the edge. It takes quite a
> bit of getting used to realize that (0,0) corresponds to infinite
> edge-lengths and it seems unnecessary with a little work on the part
> of the authors to do the conversions that will make the figs more
> natural for readers.

%das: @EM how do you feel about just plotting 1-x and 1-y and explaining that this is intuitively more like branch lengths?

### 2018-05-09 joint_inf.tex
remove farris from plot
answering EMs
more clarifications

Fidelities have useful algebraic properties.
As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).

@@

%das: TODO, rerun this figure through inkscape with updated title
\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

%EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the InvFels tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
\[
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
\]
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
\[
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
\]
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\[
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
\]
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

>>>>

Fidelities have useful algebraic properties.
As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
We show in the Appendix that, for the InvFels tree, the maximum likelihood fidelities require $\theta_1=\theta_3$ and $\theta_2=\theta_4$, so we subsequently fix fidelity parameters to $t=\{x,y,x,y,w\}$.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

%EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
%das: fixed
We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the InvFels tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
\[
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
\]
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
\[
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
\]
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\[
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
\]
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
